---
title: "A3: Incarceration"
output: html_document
---

### Introduction

The United States prison system shows racial disparities influenced by various social and political factors. This report delves into incarceration trends using data collected by the Vera Institute, focusing on variables such as urbanicity, ethnicity, and gender. The aim is to uncover patterns that highlight how incarceration rates vary across different demographics and regions, particularly how these rates differ in rural versus urban areas.

### Summary Information

The analysis of the dataset reveals several significant statistics:

- In 2018, the average jail population rates for females and males were 'r round(avg_female_jail_pop_rate, 2)' and 'r round(avg_male_jail_pop_rate, 2)' per 100,000 people, respectively.

- The total number of POC prisoners was highest in 'r paste(highest_poc_county$county_name, highest_poc_county$state)' with ’r paste(highest_poc_county$poc_prisoners)‘ prisoners.

- The year 2009 had the highest recorded jail population rate for any county, with 'r paste(highest_year_county$county_name)' county reaching 'r paste(highest_year_county$total_jail_pop_rate)' per 100,000 people.

### The Dataset

Who collected the data?  
The data was collected by the Vera Institute of Justice.

How was the data collected or generated?  
The data was generated by combining two Bureau of Justice Statistics (BJS) collections: the Census of Jails, conducted every five to eight years since 1970, and the Annual Survey of Jails , conducted annually in non-census years since 1982.

Why was the data collected?  
The data was collected to address the overcriminalization and mass incarceration of people of color, immigrants, and individuals experiencing poverty, aiming to transform the justice system.

How many observations (rows) are in your data?  
There are 153,811 observations in the dataset.

How many features (columns) are in the data?  
There are 23 features in the dataset.

What, if any, ethical questions or questions of power do you need to consider when working with this data?  
It is important to consider the biases that may exist due to varying reporting practices and missing data, particularly from smaller counties and earlier years.

What are possible limitations or problems with this data?   (at least 200 words)
The data has several limitations, including significant missing values that could misrepresent trends. Inconsistent reporting practices and the aggregation of racial subgroups might mask specific disparities. The dataset does not account for policy changes, socio-economic factors, or criminal justice reforms that influence incarceration rates, making it difficult to attribute trends to specific causes

### Trends Over Time Chart

Trends Over Time Chart displays the disparity in incarceration rates among different racial groups in America from 1990 to 2018. Black Americans have consistently faced higher incarceration rates, though there has been some improvement since the late 1990s. Meanwhile, White Americans have seen a steady increase in incarceration rates, whereas other racial groups have generally experienced stagnation or decline. These trends emphasize the need for continued attention to racial disparities in the criminal justice system.


``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

This chart examines how urbanization impacts incarceration rates among different racial groups. As America develops, it's crucial to understand how gentrification and urban growth influence these trends. The data shows that higher urbanicity generally correlates with lower incarceration rates across most ethnic groups, though Black Americans continue to face disproportionately high incarceration rates.


``````{r, echo = FALSE, code = readLines("chart2_example.R")}
```

### Map

This map highlights how urbanization affects incarceration rates in each state. Generally, states with larger populations see increased incarceration rates with higher urbanization. Utah is an outlier, likely due to data anomalies. Additionally, states along the Gulf of Mexico show significantly higher changes in incarceration rates, potentially due to stricter immigration policies.


``````{r, echo = FALSE, code = readLines("map_example.R")}
```
