---
title: "A3: Incarceration"
output: html_document
---

### Introduction

The United States prison system shows racial disparities influenced by various social and political factors. This report delves into incarceration trends using data collected by the Vera Institute, focusing on variables such as urbanicity, ethnicity, and gender. The aim is to uncover patterns that highlight how incarceration rates vary across different demographics and regions, particularly how these rates differ in rural versus urban areas.

### Summary Information

The analysis of the dataset reveals several significant statistics:

- In 2018, the average jail population rates for females and males were 'r round(avg_female_jail_pop_rate, 2)' and 'r round(avg_male_jail_pop_rate, 2)' per 100,000 people, respectively.

- The total number of POC prisoners was highest in 'r paste(highest_poc_county$county_name, highest_poc_county$state)' with ’r paste(highest_poc_county$poc_prisoners)‘ prisoners.

- The year 2009 had the highest recorded jail population rate for any county, with 'r paste(highest_year_county$county_name)' county reaching 'r paste(highest_year_county$total_jail_pop_rate)' per 100,000 people.

### The Dataset

Who collected the data?  
The data was collected by the Vera Institute of Justice.

How was the data collected or generated?  
The data was generated by combining two Bureau of Justice Statistics (BJS) collections: the Census of Jails, conducted every five to eight years since 1970, and the Annual Survey of Jails , conducted annually in non-census years since 1982.

Why was the data collected?  
The data was collected to address the overcriminalization and mass incarceration of people of color, immigrants, and individuals experiencing poverty, aiming to transform the justice system.

How many observations (rows) are in your data?  
There are 153,811 observations in the dataset.

How many features (columns) are in the data?  
There are 23 features in the dataset.

What, if any, ethical questions or questions of power do you need to consider when working with this data?  
It is important to consider the biases that may exist due to varying reporting practices and missing data, particularly from smaller counties and earlier years.

What are possible limitations or problems with this data?   (at least 200 words)
The data has several limitations, including significant missing values that could misrepresent trends. Inconsistent reporting practices and the aggregation of racial subgroups might mask specific disparities. The dataset does not account for policy changes, socio-economic factors, or criminal justice reforms that influence incarceration rates, making it difficult to attribute trends to specific causes

### Trends Over Time Chart

This chart highlights a significant issue with incarceration rates in America by displaying the incarceration rates of different racial groups from 1990 to 2018. The data shows that Black Americans have consistently been incarcerated at much higher rates compared to other groups.Another notable trend is the steady increase in incarceration rates for White Americans.


``````{r, echo = FALSE, code = readLines("chart1_example.R")}
```

### Variable Comparison Chart

This chart compares the jail population rates of white individuals to those of black individuals in King County from 1990 to 2018. The chart aims to explore whether there is a correlation between the incarceration rates of black and white individuals to uncover any racial disparities in incarceration trends. The analysis reveals that there is no strong correlation between the two groups, with a correlation coefficient of 0.01726, indicating that the incarceration trends of black individuals do not significantly impact the incarceration trends of white individuals and vice versa.


``````{r, echo = FALSE, code = readLines("chart2_example.R")}
```

### Map

This map illustrates how state incarceration rates are influenced by urbanization. An interesting pattern observed is that states with larger populations tend to see a positive change in incarceration rates as urbanization increases. However, some anomalies, like Utah, stand out due to data outliers. Additionally, states bordering the Gulf of Mexico exhibit significantly higher changes in incarceration rates across different urbanicity levels, possibly due to stricter immigration policies leading to higher incarceration rates in these states.


``````{r, echo = FALSE, code = readLines("map_example.R")}
```
